Is essentiality of genes conserved in Enterobacteriaceae?

### 2017-02-13 documents/manuscript.tex
Edited manuscript, found some interesting genes that are essential in some clades

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

@@

% this is required to include graphics
\usepackage{graphicx}
\graphicspath{{../figures/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar

@@

\section*{Abstract}
Many genes have been identified with advances in sequencing technology and genome annotation methods. However, not all of these genes are of the same importance. We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae. We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes. After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level. Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.

% now start line numbers
%\linenumbers

% the * after section prevents numbering
\section{Introduction}
With the advent of sequencing technologies and genome annotation methods many genes have been identified. However, not all of these genes are of the same importance for the growth of an organism. So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}. These studies can give us new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising minimal genomes \cite{hutchison_global_1999, hutchison_design_2016, reus_blueprint_2016}. Different methods have been used for studying the essentility of genes in prokaryotes. Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}. Another method that is widely used due to its simplicity is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}. In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}. This can lead to the identification of essential genes.

Now that the essentiality of genes can be evaluated using different methods, it is possible to compare the essentiality data in different organisms and investigate the differentiation of essentiality of their genes. Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism. Freed et al.\@ \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that some genes have gained essentiality in \textit{Shigella flexneri} while there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}. Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi and found that the essentiality of genes differs in different organisms. In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges. Although there are many studies on differentiation of essentiality in different organisms, these studies usually include two or three strains.

Our aim is to study the essentiality of genes in an evolutionary framework in 14 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}). Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

\begin{figure}
\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in EcoGene \cite{zhou_ecogene_2013}. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
\label{fig:species-tree}
\end{figure}

\section{Results and discussion}
Throughout time, species can gain or lose genes. We investigated if these gene gain and losses are related to the essentiality of the genes. In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.

\subsection{Are there biases in transposon mutagenesis data?}
To evaluate the essentiality of a gene, the number of insertions within that gene was measured as explained in~\ref{sec:essentiality}. However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis. Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}. We performed a detailed study of these biases. The biases that we have studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.

%\subsection{Length bias}
%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions. As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes. 
%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes. It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
%\begin{figure*}
%\centering
%\includegraphics[page=42, scale=0.3]{biases.pdf}
%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
%\label{fig:length-bias}
%\end{figure*}

\subsubsection{Origin of replication bias}
One possible source of bias is the distance from origin of replication. When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away. This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions. The other factor that can affect the results is if essential genes are clustered near the origin by nature. Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication. However, the essential genes are more frequent in the leading strand than the lagging one.

To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}. The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.

\begin{figure*}
\centering
\begin{tabular}{c c c}
\includegraphics[page=2, scale=0.25]{biases.pdf}&
\includegraphics[page=5, scale=0.25]{biases.pdf}&
\includegraphics[page=8, scale=0.25]{biases.pdf}\\
\includegraphics[page=11, scale=0.25]{biases.pdf}&
\includegraphics[page=14, scale=0.25]{biases.pdf}&
\includegraphics[page=17, scale=0.25]{biases.pdf}\\
\includegraphics[page=20, scale=0.25]{biases.pdf}&
\includegraphics[page=23, scale=0.25]{biases.pdf}&
\includegraphics[page=26, scale=0.25]{biases.pdf}\\
\includegraphics[page=29, scale=0.25]{biases.pdf}&
\includegraphics[page=32, scale=0.25]{biases.pdf}&
\includegraphics[page=35, scale=0.25]{biases.pdf}\\
&\includegraphics[page=38, scale=0.25]{biases.pdf}&\\
\end{tabular}
\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes. The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance. The red curves show the loess curve when the smoothness parameter is $0.2$.}
\label{fig:distance-bias}
\end{figure*}

\subsubsection{Preferred insertion motif bias}
Another concern while inferring essentiality from transposon mutagenesis data is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, instead of their essentiality level. For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome. The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.

In addition, we investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index (Fig.\@  \ref{fig:GC-bias}). As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant. A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions \cite{kimura_nucleoid_2016}. The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (Fig.\@ \ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.

\begin{itemize}
\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
\end{itemize}

\begin{figure*}
\begin{tabular}{c c}
\includegraphics[scale=0.55]{100logo-bits.pdf}&
\includegraphics[scale=0.55]{100logo-prob.pdf}
\end{tabular}
\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome. On the left the height of each character corresponds to a bit score for that character (i.e. $2-\sum f_a\times\log_2f_a-\frac{1}{\ln2}\times\frac{3}{2\times n}$, where $f_a$ is the relative frequency of base $a$ and $n$ is the number of sequences). To put it in simple words, the height of the set of characters shows how biased that position is and the height of each character shows the amount of bias towards that character. On the right the height of each character shows the relative frequency of that character.}
\label{fig:logos}
\end{figure*}

\begin{figure*}
\centering
\begin{tabular}{c c c}
\includegraphics[page=1, scale=0.25]{biases.pdf}&
\includegraphics[page=4, scale=0.25]{biases.pdf}&
\includegraphics[page=7, scale=0.25]{biases.pdf}\\
\includegraphics[page=10, scale=0.25]{biases.pdf}&
\includegraphics[page=13, scale=0.25]{biases.pdf}&
\includegraphics[page=16, scale=0.25]{biases.pdf}\\
\includegraphics[page=19, scale=0.25]{biases.pdf}&
\includegraphics[page=22, scale=0.25]{biases.pdf}&
\includegraphics[page=25, scale=0.25]{biases.pdf}\\
\includegraphics[page=28, scale=0.25]{biases.pdf}&
\includegraphics[page=31, scale=0.25]{biases.pdf}&
\includegraphics[page=34, scale=0.25]{biases.pdf}\\
\includegraphics[page=37, scale=0.25]{biases.pdf}&&
\includegraphics[page=40, scale=0.25]{biases.pdf}\\
\end{tabular}
\caption{The plots show the ratio of G-C bases in the genes normalised by the lengths of the genes against their insertion indices. The red curves show the loess curve when the smoothness parameter is $0.2$.}
\label{fig:GC-bias}
\end{figure*}

\begin{figure*}
\includegraphics[scale=0.55]{lowgc-pval.pdf}
\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level. The red line shows P-value = 0.05. The P-values have been calculated using Fisher's exact test and corrected using Benjamini-Hochberg-Yekutieli.}
\label{fig:gc-pval}
\end{figure*}

\subsubsection{Positional bias within genes}
The other question that we tried to answer was whether insertions are tolerated in some regions in a gene. For example, can essential genes tolerate insertions at their 3' end without losing their funtionality? To address this question, we divided every gene into percentiles and calculated the mean insertion index for each percentile. Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location when considering all genes together. We also studied the bias in three of the groups defined in Section~\ref{sec:essentiality}: essential genes, non-essential genes, and beneficial losses. The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses. The case for the non-essential genes is similar to all genes. High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions. On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors. \textcolor{red}{\{To be tested\}}

\begin{figure*}
\begin{tabular}{c c}
\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
\end{tabular}
\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right). The genes are divided into 3 segments: 5\% of
the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
\label{fig:insertion-position-bias}
\end{figure*}

%\begin{itemize}
%\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
%\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
%\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
%\item The number of insertions on the 3' and 5' ends is more than the internal region in essential genes and less than the internal region in beneficial losses. (Fig.~\ref{fig:insertion-position-bias})
%\end{itemize}
\subsection{Essentiality and conservation}
Essential genes are needed for the growth of organisms. Because of that, one might think that essential genes should not be lost in a short period of time throughout evolution, unless they are no longer needed in new organisms or they are replaced by new pathways. Therefore, it is expected that most of the essential genes are conserved in different organisms from the same family. We have tested this idea by comparing the essentiality and conservation of genes in Enterobacteriaceae family.

\subsubsection{Gene classes}
%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality. Plenty of methods are proposed for this purpose. Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall. We have used Hieranoid to cluster the sets of orthologous genes. In addition, we intended to study the essentiality of genes in paralogous genes. For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes. For this, we divided the genes into different levels of essentiality (essential genes, ambiguous, non-essential genes, beneficial losses). We first normalised the biases that exist in the data using the method explained in Section~\ref{sec:biascorrection} and then defined the essentiality level of genes using the method explained in Section~\ref{sec:essentiality}. Moreover, we grouped the genes into different classes of conservation (genus specific, single copy, multi-copy) using the method explaind in Section~\ref{sec:conservationclasses}.

The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}. The high number of single copy genes in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality. However, the relatively high number of essential genes in genus specific class implies that each genus has a set of essential genes that makes it distinct. The figure also shows that beneficial losses are over represented in genus specific class. Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run. Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential. This can be explained by the redundancy that genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}. In the presence of two redundant variations of one gene, if we knock out one copy by transposon mutagenesis, the other copy compensates and the organism can still survive.

\begin{figure}
\centering
\includegraphics[scale=0.5]{cluster-essentiality.pdf}
\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes. Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters. The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
\label{fig:iidist}
\end{figure}

>>>>

% adjust caption style
\usepackage[font=small,aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

@@

% this is required to include graphics
\usepackage{graphicx}
\usepackage{subcaption}
\graphicspath{{../figures/}}

@@

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar
\usepackage{amsmath}

@@

\section*{Abstract}
%Many genes have been identified with advances in sequencing technology and genome annotation methods. However, not all of these genes are of the same importance. We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae. We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes. After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level. Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.

% now start line numbers
\linenumbers

% the * after section prevents numbering
\section{Introduction}
%The study of cells can give us new insights for understanding the foundation of life. However, the naturally existing cells are very complex which hampers the study of their constituents. Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.

Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions. Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes. Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}. Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}. Infectious diseases are among the top major causes of mortality worldwide. Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}. Therefore, it is urgent to find new drugs for infectious diseases. New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease. Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]

In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}. With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}. Therefore, the use of experimental methods for the identification of essential genes is vital. Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods. Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12. The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes. Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}. In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential. Both of these methods are labour intensive and are dependent on the accuracy of genome annotations. Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}. These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}. Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis. After a growth phase, mutants that are fitter outnumber the less fit ones. Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not. A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.

There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms. Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}. On the other hand, accessory essential genes are helpful in the study of specific lineages. Accessory essential genes may be useful in species-specific antibiotic discovery. Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.

%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
Freed et al. \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}. These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and

%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes. Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}. For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential. Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}. These include genes involved in replication and genes related to ribosome and its accessory proteins. The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S. typhi} and 10 genes vice versa. Barquist et al. \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113. 
%THIS IS IMPORTANT:
These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis. Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors. Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.

Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes. We are going to extend these studies to a larger scale in Enterobacteriaceae family. Our aim is to study the essentiality of genes in in 16 different organisms. These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655. The phylogenetic tree containing all these species is depicted in Fig.\@ \ref{fig:species-tree}.

%WHAT FATEMEH DID....

%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}. In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study. In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential. We first performed a detailed study of biases than can influence the inference of essentiality. Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome). We have also investigated how essentiality changes in the phylogenetic tree for these organisms.

\begin{figure}
\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
\caption{The species tree containing the 16 genomes used in this study. We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
\label{fig:species-tree}
\end{figure}

\section{Results and Discussion}
\subsection{Incorporating multiple measures of gene essentiality improves prediction}
In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
\subsubsection{Comparison of old methods}
A number of methods have been used for evaluating the essentiality of genes using transposon insertion data. Freed et al. \cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes. These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length. Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.

Barquist et al. \cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes. Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential). They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.

Turner et al. \cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.

We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling. To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}. The results are depicted in Figure~\ref{fig:benchmark}. Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods. Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
\subsubsection{PCA based method}\label{sec:essentiality-call}
We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region. These predictors were selected as they were more accurate and less dependent on each other. Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component. As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods. We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}. Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation. We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper. The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential. The rest are non-essential genes. The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95. Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.

\begin{figure}
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=1, scale=0.4]{essential-call-comparison-ETEC.pdf}
  \caption{}
  \label{fig:benchmark}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[page=10, scale=0.4]{essential-call-comparison-ETEC.pdf}
  \caption{}
  \label{fig:pca}
\end{subfigure}
\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is. (b) The distribution for our proposed method (NPEQ)}
\label{fig:essentiality-call}
\end{figure}

%\begin{figure*}
%\centering
%\begin{tabular}{c c c}
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
%\end{tabular}
%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
%\label{fig:benchmark}
%\end{figure*}
